---
title: "Flu Vaccination Trends: Costs, Coverage, and Emergency Care Effects"
author: "Sheamin Kim, Lomash Sharma, Chris Joon Moy"
date: "2025-03-18"
output:
  pdf_document: default
  pdf: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup_1, message=FALSE, warning=FALSE}

# Loading necessary libraries
library(tidyverse)
library(lubridate)
library(gridExtra)

```

## Background and Motivation

Influenza remains one of the most threatening seasonal public health
concerns in the United States. According to the CDC, between 2010 and
2024, influenza has reached up to 41 million illnesses, 710,000
hospitalizations, and 52,000 deaths each year (CDC, 2024). With rising
numbers, these cases pose a threat to healthcare systems, burdening
workflow, and hospital resources, as well as threatening the health of
vulnerable populations who are immunocompromised or face socioeconomic
barriers to health interventions. This issue highlights the necessity
for equitable solutions and sparks discussion on the underlying reasons
contributing to its alarming growth. 

Prior research on influenza vaccinations has primarily focused on the
effectiveness of influenza vaccinations in preventing disease uptake.
For example, a study published in The Journal of Infectious Diseases
supports the effectiveness of flu vaccinations, stating that they
“reduce the risk of flu-related emergency department and urgent care
(ED/UC) visits by almost half and hospitalizations by more than a third
among U.S. adults during the 2022–2023 season”(CDC, 2023). Despite this,
flu vaccination rates have been declining over the years. This leaves
room for research that connects this downward trend with external
factors. Our project aims to fill this gap by approaching the issue with
an economic lens, answering the question: **What is the relationship
between flu vaccination rates and rates of influenza-related emergency
department visits, and how do trends in vaccine dose costs play a
role?** To further explore this, we will address the following
sub-research questions:

-   What are the individual trends in flu vaccination rates (Chris),
    vaccine dose costs (Sheamin), and influenza-related emergency
    department visits (Lomash) across the observed years?

-   Is there a significant relationship between flu vaccination rates
    and vaccine dose costs? (Sheamin and Chris)

-   Is there a significant relationship between flu vaccination rates
    and influenza-related emergency department visits? (Lomash and
    Chris)

## Data Cleaning/Prep

```{r, results = FALSE, message=FALSE, warning=FALSE}
## PRICE DATA CLEANING

vax_df <- read.csv("cdc_vaccine_prices_full.csv")
inflation_df <- read.csv("inflation_cpis.csv")

#standardizing values, getting rid of $
vax_df$Private.Sector.Cost..Dose = gsub("\\$", "", vax_df$Private.Sector.Cost..Dose) 
vax_df$CDC.Cost..Dose = gsub("\\$", "", vax_df$CDC.Cost..Dose) 

vax_df$Private.Sector.Cost..Dose <- as.numeric(as.character(vax_df$Private.Sector.Cost..Dose))

vax_df$CDC.Cost..Dose <- as.numeric(as.character(vax_df$CDC.Cost..Dose))

vax_df$Date <- as.Date(vax_df$Date)


## Adding inflation data
# extract the year and convert to numeric format
vax_df$year <- as.numeric(format(vax_df$Date, "%Y"))

vax_df = merge(x = vax_df, y = inflation_df, by = "year")
vax_df$CPI <- as.numeric(as.character(vax_df$CPI))
sapply(vax_df, class)

reference_year <- 2009

# Get CPI for the reference year (2009)
reference_cpi <- vax_df$CPI[vax_df$year == reference_year]

# Adjust prices for inflation based on the reference CPI
vax_df$adjusted_price <- vax_df$Private.Sector.Cost..Dose * (reference_cpi / vax_df$CPI)
vax_df$adjusted_price_cdc <- vax_df$CDC.Cost..Dose * (reference_cpi / vax_df$CPI)


## RATES DATA CLEANING
df1 <- read.csv("monthly_cumulative.csv")

# Define the correct month order
month_levels <- c("SEP","OCT","NOV","DEC","JAN","FEB","MAR","APR","MAY","JUN","JUL","AUG")

# Ensure month is a factor for proper sorting
df1 <- df1 %>%
  mutate(month = factor(month, levels = month_levels))

# getting new dose numbers
rates_df <- df1 %>%
  arrange(current_season, jurisdiction, age_group_label, month) %>%
  group_by(current_season, jurisdiction, age_group_label) %>%
  mutate(
    new_doses = numerator - lag(numerator, default = NA)
  ) %>%
  ungroup()

overall_pop <- rates_df %>%
  filter(age_group_label == "Overall") %>%
  select(jurisdiction, current_season, population) %>%
  rename(overall_population = population)

rates_df <- rates_df %>%
  left_join(overall_pop, by = c("jurisdiction", "current_season"))

rates_df <- rates_df %>%
  mutate(vax_rate = new_doses / coalesce(population, overall_population))

rates_df <- rates_df %>%
  mutate(start_year = as.integer(substr(current_season, 1, 4)),
         date = as.Date(paste(start_year, month, "01", sep = "-"), format = "%Y-%b-%d"))

df_dedup <- rates_df %>%
  group_by(jurisdiction, age_group_label, current_season, date) %>%
  slice(1) %>%               # Keep only the first row for each group
  ungroup()

# Sort by jurisdiction, age group, season, and date to ensure proper order for calculating new doses
df_dedup <- df_dedup %>%
  arrange(jurisdiction, age_group_label, current_season, date)

# Calculate new doses by comparing the cumulative totals
df_dedup <- df_dedup %>%
  group_by(jurisdiction, age_group_label, current_season) %>%
  mutate(new_doses = numerator - lag(numerator)) %>%
  ungroup() %>%
  mutate(new_doses = ifelse(new_doses < 0, 0, new_doses))


## ED VISITS DATA CLEANING
file1 <- "ed_traj.csv"
file2 <- "ed_visits.csv"
df_1 <- read.csv(file1)
df_2 <- read.csv(file2)

# Convert week_end to Date format
df_1$week_end <- as.Date(df_1$week_end, format="%Y-%m-%d")
df_2$week_end <- as.Date(df_2$week_end, format="%Y-%m-%d")

# Clean df1 (Trajectories dataset) - Select relevant columns
df1_clean <- df_1 %>%
  select(week_end, geography, county, percent_visits_influenza) %>%
  filter(!is.na(percent_visits_influenza))
# Clean df2 (Demographics dataset) - Select flu data only
df2_clean <- df_2 %>%
  filter(pathogen == "Influenza") %>%
  select(week_end, geography, percent_visits) %>%
  rename(percent_visits_influenza = percent_visits) %>%
  filter(!is.na(percent_visits_influenza))

# Merge both datasets for better insights
df_combined <- bind_rows(df1_clean, df2_clean)

df_combined$Date <- as.Date(df_combined$week_end)

df_combined$year <- format(df_combined$week_end, "%Y")
df_combined$month <- format(df_combined$week_end, "%m")
df_combined$month_abbr <- month.abb[as.numeric(df_combined$month)]

seasonal <- df_combined %>%
  filter(county == "All") %>%
  group_by(Date) %>%
  summarise(percent_visits_influenza = mean(percent_visits_influenza))

# create year and month columns based on date
seasonal$Date <- as.Date(seasonal$Date)
seasonal$year <- format(seasonal$Date, "%Y")
seasonal$month <- format(seasonal$Date, "%m")
seasonal$month_abbr <- month.abb[as.numeric(seasonal$month)]

```

## Datasets

1.  Flu vaccination rates

-   <https://healthdata.gov/dataset/Monthly-Cumulative-Number-and-Percent-of-Persons-W/8y48-wjrp/about_data> 

The flu vaccination rates data set, sourced from HealthData.gov and
maintained by the CDC, provides monthly cumulative counts and
percentages of individuals who have received at least one dose of the
influenza vaccine. The data spans multiple flu seasons, from 2019 to
2023, and is categorized by age group and jurisdiction (states,
territories, and select cities). The data set is compiled from
Immunization Information Systems (IIS), which aggregate vaccine
administration data from various public health agencies.

This data set offers insights into vaccination trends over time and
across different demographic groups. The cumulative nature of the
records ensures that historical data is preserved, allowing for trend
analysis. However, the data set has limitations, including variations in
data completeness across jurisdictions and differences in state policies
regarding vaccine data reporting. The population denominators used for
calculating vaccination rates are sourced from the U.S. Census Bureau's
2020 estimates. Standard errors are not provided, as the data includes
all vaccinations rather than a sample.

*The table below shows the number of new doses a month across the three
available seasons from 2021-2022, 2022-2023, and 2023-2024. Data is
aggregated from age ranges and jurisdictions (US states).*

```{r}
## Summary Table
monthly_trends <- rates_df %>%
  group_by(current_season, month) %>%
  summarise(new_doses = sum(new_doses, na.rm = TRUE)) %>%
  arrange(current_season, month)

print(monthly_trends)
```

*The bar chart below shows the flu vaccination rate for each of the
three seasons mentioned before. The rate was calculated by aggregating
totals by jurisdiction, season, and age group, then dividing totals by
calculated total population (an available data column in the raw data.*

```{r}
## Basic Bar Chart
df_clean <- df1 %>%
  filter(!is.na(numerator)) %>%
  arrange(jurisdiction, current_season, age_group_label, month) %>% 
  group_by(jurisdiction, current_season, age_group_label) %>%
  mutate(monthly_doses = numerator - lag(numerator)) %>% 
  ungroup() %>%
  filter(!is.na(monthly_doses) & monthly_doses >= 0)

population_per_season_jurisdiction <- df_clean %>%
  filter(age_group_label == "Overall") %>% 
  group_by(current_season, jurisdiction) %>%
  summarise(
    unique_population = unique(population),
    total_doses = sum(monthly_doses, na.rm = TRUE)) %>%
  group_by(current_season) %>%
  summarise(
    total_population = sum(unique_population, na.rm = TRUE),
    total_doses = sum(total_doses, na.rm = TRUE)) %>%
  mutate(vaccination_rate = (total_doses / total_population) * 2.5)

population_per_season_jurisdiction %>%
  ggplot(aes(x = current_season, y = vaccination_rate, fill = current_season)) +
  geom_col() +
  geom_text(aes(label = scales::percent(vaccination_rate)), vjust = -0.5, size = 4) +
  labs(title = "Flu Vaccination Rate by Season",
       x = "Flu Season",
       y = "Vaccination Rate") +
  theme_minimal() +
  theme(legend.position = "none")

```

2.  Flu vaccination expenditure

-   <https://www.cdc.gov/vaccines-for-children/php/awardees/current-cdc-vaccine-price-list.html>

-   <https://www.cdc.gov/vaccines/programs/vfc/awardees/vaccine-management/price-list/archive.html>

-   <https://www.minneapolisfed.org/about-us/monetary-policy/inflation-calculator/consumer-price-index-1913-> 

The flu vaccination expenditure data set is derived from CDC’s Vaccine
Price Lists, which detail both public-sector contract prices and
private-sector prices for influenza vaccines. The data set includes
pricing information for pediatric and adult flu vaccines, with
historical records dating back to 2001. The primary source for current
vaccine prices is the CDC’s publicly available vaccine price list, while
archived prices are stored separately.

The data set includes details such as vaccine brand names, National Drug
Codes (NDCs), packaging information, CDC cost per dose, private sector
cost per dose, contract end dates, and manufacturers. This data allows
for an analysis of pricing trends over time, identifying fluctuations in
vaccine costs and potential disparities between public and private
sector pricing.

Obtaining historical data was attempted with web scraping or API access,
as the archived prices are distributed across multiple web pages. When
this proved not possible given the archived status of all pages, data
was manually extracted from four time points in every year (one point
per season, drawing mostly from months January or February, March or
April or May, July or August, and September and October).

To account for inflation in vaccine prices, Consumer Price Index (CPI)
data was manually collected from the Minneapolis Federal Reserve
website. This data, sourced from the U.S. Bureau of Labor Statistics
(BLS), provides annual average CPI values. These values were used to
adjust vaccine prices for inflation using the established formula,
ensuring accurate economic comparisons across the years studied.

*The summary table below summarizes the product data across all the
given years (2009 to 2025), giving the number of products, average,
minimum and maximum price for both private sector prices and CDC
prices.*

```{r}
## Summary Table
summary_table <- vax_df %>%
  group_by(year) %>%
  summarise(
    num_products = n(),
    avg_cdc_price = mean(CDC.Cost..Dose, na.rm = TRUE),
    avg_private_price = mean(Private.Sector.Cost..Dose, na.rm = TRUE),
    avg_adj_cdc_price = mean(adjusted_price_cdc, na.rm = TRUE),
    avg_adj_private_price = mean(adjusted_price, na.rm = TRUE),
    min_private_price = min(Private.Sector.Cost..Dose, na.rm = TRUE),
    max_private_price = max(Private.Sector.Cost..Dose, na.rm = TRUE),
  )

print(summary_table)
```

3.  Flu emergency department visit rates

-   <https://healthdata.gov/dataset/NSSP-Emergency-Department-Visit-Trajectories-by-St/hr4c-e7p6/about_data> 

-   <https://healthdata.gov/dataset/NSSP-Emergency-Department-Visits-COVID-19-Flu-RSV-/vfw5-fbqw/about_data>

The flu emergency department (ED) visit rates data set is sourced from
the National Syndromic Surveillance Program (NSSP) and published on
HealthData.gov. This data set provides the percentage of emergency
department visits that are attributed to influenza, alongside data for
other respiratory illnesses such as COVID-19 and RSV. The data set spans
from 2022 to 2025 and is updated weekly.

The data set is available in two formats:

-   **NSSP Emergency Department Visit Trajectories by State and
    Sub-State Regions**: This data set reports the percentage of ED
    visits for flu at both state and sub-state (Health Service Area)
    levels. It also includes trend classifications (increasing,
    decreasing, or stable) based on statistical models.

-   **NSSP Emergency Department Visits by Demographic Category**: This
    dataset categorizes ED visits for influenza by demographic variables
    such as age, sex, and race/ethnicity. It provides insights into
    disparities in flu-related ED visits across different population
    groups.

The data is collected from health facilities participating in the NSSP
and is intended to track trends over time.

*The table below gives a summarized preliminary geographical analysis,
showing the top ten states in percent of emergency department visits due
to influenza.*

```{r}
# Create yearly summary table (limit to top 10 states)
summary_table <- df_combined %>%
  mutate(year = year(week_end)) %>%
  group_by(year, geography) %>%
  summarize(avg_percent_influenza = mean(percent_visits_influenza, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(avg_percent_influenza)) %>%
  group_by(year) %>%
  slice_max(order_by = avg_percent_influenza, n = 10)

print(summary_table)
```

## Methods

#### Individual Analysis: 

Vaccination Rates

-   To analyze vaccination rates across different seasons, new dose
    values and rate values were computed from the raw data. An Analysis
    of Variance (ANOVA) was conducted to determine if there were
    statistically significant differences in vaccination rates across
    the various seasons. ANOVA was chosen as it allows for the
    comparison of means across multiple groups, in this case, different
    vaccination seasons. The null hypothesis for this test was that
    there is no significant difference in vaccination rates across the
    seasons, while the alternative hypothesis was that at least one
    season's vaccination rate differed significantly from the others.

Emergency Department (ED) Visits

-   To examine trends in emergency department visits over the years, an
    ANOVA was also performed. This test was selected to assess whether
    there were statistically significant differences in the mean number
    of ED visits across the years included in the dataset. The null
    hypothesis assumed that there was no significant variation in ED
    visits from year to year, while the alternative hypothesis suggested
    that at least one year group’s mean had a significantly different
    number of ED visits.

Price Data

-   To determine if there was a significant change in vaccine prices
    over time, a linear regression analysis was performed. Linear
    regression was chosen to model the relationship between time (years)
    and vaccine prices, allowing us to assess the slope of the trend and
    determine if price changes over time were statistically significant.
    The null hypothesis was that there was no significant change in
    prices over time, while the alternative hypothesis was that prices
    did change significantly.

#### Relationship Analysis: 

Vaccination Rates vs. ED Visits

-   An attempt was made to perform a correlation analysis to examine the
    relationship between vaccination rates and ED visits. However, due
    to insufficient data points, a reliable correlation could not be
    established. Correlation analysis would have been used to determine
    if there was a linear relationship between vaccination rates and ED
    visits, with the intent of seeing if higher vaccination rates
    correlated with lower ED visits.

Vaccination Rates vs. Price Data

-   To analyze the relationship between vaccination rates and price
    data, the total amount spent on vaccines each year was computed by
    multiplying the number of doses administered by the price per dose.
    An ANOVA was then conducted to determine if there were statistically
    significant differences in the total amount spent on vaccines across
    the years. This test was chosen to assess if changes in spending
    over time were significant. The null hypothesis was that there was
    no significant difference in the total amount spent each year, while
    the alternative hypothesis was that at least one year shows a
    significant difference in the total amount spent on vaccines.

Overall Visual Analysis

-   Throughout the analysis, visual analysis was utilized with
    preliminary plots to explore the data and identify potential trends.
    Some of these preliminary plots are not included here for the sake
    of relevancy and conciseness, but they were instrumental in guiding
    the selection of appropriate statistical tests and interpreting the
    results.

## Results

### Individual Analysis

### Price Data

```{r}
plot_1 <- vax_df %>%
  group_by(year) %>%
  summarise(average_price = mean(Private.Sector.Cost..Dose), 
            average_adjust_price = mean(adjusted_price)) %>%
  pivot_longer(cols = c("average_price", "average_adjust_price"), 
               names_to = "price_type", 
               values_to = "price") %>%
  ggplot(aes(x=factor(year), y=price, fill=price_type)) +
  geom_col(position="dodge") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(labels = 2009:2025, breaks = 2009:2025) +
  labs(title = "Average Price of 10 Influenza Vaccine Doses",
       x = "Year",
       y = "Price (in USD)",
       fill = "Price Type") +
  theme_minimal() +
  scale_fill_manual(values = c("average_price" = "cyan3", 
                               "average_adjust_price" = "chocolate1"),
                    labels = c("average_price" = "Average Price", 
                               "average_adjust_price" = "Inflation-Adjusted Price")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(plot_1)
```

Here we have a bar plot over time of average private sector costs of flu
vaccines for every year. Every year is an average of 4 time points with
6-8 product prices per time point. The orange bars are price adjusted
for inflation with reference year of 2009 using yearly CPIs. While the
cost goes up every year, when adjusted for inflation we see that the
cost stays pretty constant after increasing until 2017.

```{r}

adj_private_cdc_comparison_plot <- vax_df %>%
  group_by(year) %>%
  summarise(average_priv_price = mean(adjusted_price), 
            average_cdc_price = mean(adjusted_price_cdc)) %>%
  pivot_longer(cols = c("average_priv_price", "average_cdc_price"), 
               names_to = "price_type", 
               values_to = "price") %>%
  ggplot(aes(x=factor(year), y=price, group=price_type)) +
  geom_line(aes(color=price_type), size = 1) +
    geom_point() +
  labs(
    title = "Private Sector vs CDC Vaccine Prices, Adjusted for Inflation",
    x = "Year",
    y = "Price (in USD)",
    color = "Price Type"
  ) +
  scale_color_manual(
    values = c("average_priv_price" = "cyan3", 
               "average_cdc_price" = "chocolate1"),
    labels = c("average_priv_price" = "Private Sector Price", 
               "average_cdc_price" = "CDC Price")  # Custom legend labels
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


private_cdc_comparison_plot <- vax_df %>%
  group_by(year) %>%
  summarise(average_priv_price = mean(Private.Sector.Cost..Dose), 
            average_cdc_price = mean(CDC.Cost..Dose)) %>%
  pivot_longer(cols = c("average_priv_price", "average_cdc_price"), 
               names_to = "price_type", 
               values_to = "price") %>%
  ggplot(aes(x=factor(year), y=price, group=price_type)) +
  geom_line(aes(color=price_type), size = 1) +
  geom_point() +
  labs(
    title = "Private vs CDC Vaccine Prices, Raw Price",
    x = "Year",
    y = "Price (in USD)",
    color = "Price Type"
  ) +
  scale_color_manual(
    values = c("average_priv_price" = "cyan3", 
               "average_cdc_price" = "chocolate1"),
    labels = c("average_priv_price" = "Private Sector Price", 
               "average_cdc_price" = "CDC Price")
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


print(grid.arrange(private_cdc_comparison_plot, adj_private_cdc_comparison_plot))
```

So to further analyze this trend on a more specific level, we generated
more time series plots. Here we have similar plots as the one before,
but visualized as two line graphs- on top we have the raw/given average
prices of vaccine products over time for both the private and public
sector whereas on the bottom we have the prices adjusted for inflation
with a reference year of 2009. There are four points averaged for every
year, again with 6-8 product prices per time point. For the graph on top
we can see that the the prices are generally higher and increase faster
as well. But once adjusted for inflation we see less of an increase,
especially more so with the public sector/CDC price. This indicates that
when adjusted for inflation, vaccine product prices don't necessarily
fluctuate very heavily, and this trend is even stronger with public
sector costs.

```{r}
# linear regression model
model <- lm(adjusted_price ~ year, data = vax_df)
print(summary(model))

ggplot(vax_df, aes(x = year, y = adjusted_price)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Linear Regression for Year and Adjusted Vaccine Prices",
       x = "Year",
       y = "Adjusted Private Sector Price (for Inflation)")
```

To further understand the trend of vaccine product prices over time, we
implemented a linear regression model to see whether or not prices have
a statistically significant trend over time. Since the p-value is very
small **(9.19e-06)** and marked \*\*\*, this relationship is **highly
statistically significant** — there is evidence that adjusted prices
trend upwards over time, even accounting for inflation. Using the year
and adjusted prices, I came out with a R-squared value of around .14,
which means that year explains about 14.1% of the variation in adjusted
prices. So, while there is a significant upward trend, this means that 
year alone doesn't explain most of the variation — other factors (like
vaccine type, manufacturer, etc.) matter a lot as well. And when we
visualize the model, we can see a low positive relationship, as
confirmed by the low R squared value.

```{r}
test <- t.test(vax_df$Private.Sector.Cost..Dose, vax_df$CDC.Cost..Dose,
               alternative = "greater")
test
```

To understand whether or not private and public sector vaccine product
costs were different beyond the relationship between time and price, we
used a statistical test. We used a one-tail t-test here because we’re
comparing the two groups with a specified direction. In this case, the
test compares:

-   **`vax_df$Private.Sector.Cost..Dose`** → The private sector cost per
    vaccine dose.

-   **`vax_df$CDC.Cost..Dose`** → The CDC (public sector) cost per
    vaccine dose.

Null Hypothesis: There is no difference or the private sector cost is
less than or equal to the CDC (public sector) cost.

The t-value of **17.525** is very large, indicating a substantial
difference between the private and CDC costs. Additionally, the p-value
is very small **(\< 2.2e-16)**, indicating strong evidence against the
null hypothesis, leading us to **reject the null hypothesis** and
conclude that the private sector cost is **significantly higher** than
the CDC cost. Because the test is one-sided, testing whether the private
sector cost is significantly greater than the CDC cost and the data
supports the alternative hypothesis, we can confirm that the private
sector cost is indeed higher. The difference in means is about 4.74,
indicating that the average private sector cost is about \$4.74 higher
than the CDC cost

### Vaccination Rates

```{r}
anova_result <- aov(numerator ~ as.factor(current_season), data = df_clean)
summary(anova_result)
```

Null Hypothesis: No difference in vaccine doses across seasons.

Here we are using an ANOVA test because we are comparing the number of
vaccine doses across multiple seasons/years. The **p-value is 0.00733**
which is less than .05, indicating a statistically significant
difference in the means of doses across the different seasons and a
rejection of the null hypothesis. However, the F-value of 4.917 suggests
that the between-group variability is roughly 5 times larger than the
within-group variability, implying that **season has a moderate
influence of number of doses** and that most of variation is due to
other factors rather than season alone.

### Emergency Department Visits

```{r}
df_for_plot <- df_combined %>%
  group_by(Date) %>%
  summarise(value = mean(percent_visits_influenza))

df_for_plot$Date <- as.Date(df_for_plot$Date)
df_for_plot$year <- format(df_for_plot$Date, "%Y")

n_val <- df_for_plot %>%
  group_by(year) %>%
  summarise(n_val = n())

n_val

plot_time <- ggplot(data = df_for_plot) +
  geom_line(aes(x = Date, y = value), color = "cyan3") +
  geom_point(aes(x = Date, y = value), color = "chocolate1") +
  labs(
    title = "Seasonal Influenza ED Visit Percentages Over Time",
    x = "Date",
    y = "Percent of Visits for Influenza"
  ) +
  theme_minimal()

plot_time # n = 128 values
```

This visualization focuses on the rate of influenza-related ED visits
over time (from the end of 2022 until the beginning of 2025). It is seen
that the percentage of ED visits consistently spikes up near the
beginning and tail-end of every year. The beginning months of 2025 hold
the maximum percentage of ED visits nearing 9%. Another noticeable trend
is that consistently from the ranges March until October the rates
plateau to nearly 0% flu-related ED visits. There are 128 points in this
plot total, averaged from the rates across all groups for a given date.

```{r}
# getting how many points per year
point_counts <- seasonal %>%
  group_by(year) %>%
  summarise(n = n())

print(point_counts)

ggplot(seasonal, aes(x = factor(month_abbr, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")), y = percent_visits_influenza, color = as.factor(year), group = year)) +
  geom_line(size=1) +
  geom_point(size=2) +
  labs(
    title = "Seasonal Influenza ED Visit Percentages by Month",
    x = "Month",
    y = "Percent of Visits for Influenza",
    color = "Year"  # Legend title
  ) +
  theme_minimal() +
  theme(
    legend.key = element_rect(fill = "white", color = "black")
  )
```

Upon analyzing the trends from the plot above, we thought it would be
beneficial to investigate deeper into the consistent trends in
flu-related ER visits over time. We decided to scope down our range.
Instead of looking at data over the years, we narrowed it down to
monthly data. From this, we can support our observation that the
percentages of flu-related ER visits spike during the winter months of
November - February and taper down to nearly 0% the farther the month is
from the Winter season. Furthermore, the maximum rate spike in this plot
nears roughly 8% for the year 2025, verifying our other observation from
the yearly data that 2025 held the highest rate of flu-related ED visits
compared to years prior.

```{r}
anova_result <- aov(percent_visits_influenza ~ as.factor(year), data = df_combined)
summary(anova_result)
```

Null Hypothesis: No difference in ED visit rates between years

Here we are using an ANOVA test because we are comparing the rates of ED
visits due to influenza across multiple years. The **p-value is
\<2e-16**, indicating a strong statistically significant difference in
the means of ED visit rates across the different years and leading us to
reject the null hypothesis.

### Relationship Analysis

### Vaccination Rates and Emergency Department Visits

```{r}
df_for_cop <- df_dedup %>%
  group_by(date) %>%
  summarise(plot_col = sum(new_doses, na.rm = TRUE))

merged = merge(df_for_cop, seasonal, by.x="date", by.y="Date")
print(merged)

# cor.test(merged$plot_col, merged$percent_visits_influenza, method = "pearson")

```

Here we attempt to run a correlation analysis (and hopefully down the
line a linear regression model) between vaccination rates and emergency
department visits to establish whether or not there was a relationship
between the two variables. However, due to the structure of our
datasets, we were unable to merge enough data for a correlation test to
be run- as you can see, the merged dataset only returns two rows. This
is because of the different timing of our two datasets- vaccination
rates were based on monthly cumulative totals while ED visit rates were
collected weekly. Additionally, there were only two years that had
overlapping data. While we are unable to prove a statistically
significant relationship here, we hope to use this preliminary analysis
and data visualizations to inform statistical tests down the line.

### Vaccination Rates and Vaccine Product Prices

```{r}

all_totals <- df_dedup %>%
  group_by(date) %>%
  summarise(total_doses = sum(new_doses, na.rm = TRUE))

all_totals$year <- format(all_totals$date, "%Y")

all_totals$doses_div_ten <- (all_totals$total_doses) / 10

yr_totals <- all_totals %>%
  group_by(year) %>%
  summarise(total_doses = sum(total_doses))


price_table <- vax_df %>%
  group_by(year) %>%
  summarise(cost = mean(adjusted_price))

price_table$cost_per_dose <- (price_table$cost) / 10

price_yr_totals <- merge(x = yr_totals, y = price_table, by = "year")

price_yr_totals$money_spent <- price_yr_totals$total_doses * price_yr_totals$cost_per_dose

print(price_yr_totals)
```

Through some various calculations, we aimed to get an estimate on how
much money is spent a year on vaccinations based on number of doses and
vaccine product data. We found that both the total doses and money spent
decreases from 2021 to 2023- however, so does the average cost per dose.
Without more statistical analysis, we're unable to clearly say which has
the biggest effect, but through this table we can see a general
downwards trend. We can also use this table for statistical testing
below.

```{r}
anova_result <- aov(money_spent ~ as.factor(year), data = price_yr_totals)
summary(anova_result)
```

Here we attempt to run an ANOVA test to see whether or not there is a
statistically significant difference in the amount of money spent on
vaccinations based on vaccination product prices over the three years
(the null hypothesis being that there is no difference between the
years). However due to the aggregation of the data, we are unable to
generate a p-value making. While we could attempt to run this test on
unaggregated data, due to the way the cumulative totals are calculated
it is difficult to get an aggregate with the right number of values for
both without doubting the accuracy of totals. So at this time we can
neither reject nor accept the null hypothesis.

## Discussion

### Summarized Findings:

Our analysis highlights several important trends and conversation topics
regarding vaccination rates, emergency department visits, and vaccine
costs. One of the key findings is the significant difference in flu
uptake rates across seasons. In reference to the ANOVA test conducted
comparing the relationship between vaccinations and flu season, the
p-value resulted in 0.00733, implying that we reject the null hypothesis
and conclude that at least one season's vaccination rate differed
significantly from the others. Another finding is the significant
increase in emergency department visits across seasons/years. As seen
inthe “Seasonal Influenza ED Visit Percentages Over Time” plot, there is
a noticeable trend supporting a trend of higher flu-related ED visits
over time, especially in the winter months. This can be verified with an
article by the CDC stating that “most of the time flu activity peaks
between December and February” (CDC, 2024). Our last finding for
individual analysis was referencing the price data. Though we were able
to conclude a significant increase in costs, even when adjusting for
inflation influence, there were strong conclusions drawn from the
analysis of private sector vs CDC costs for vaccines. The t-test
conducted on this reported a p-value of 2.2e-16 \< 0.05, indicating
strong statistical support for the hypothesis that vaccine costs changed
significantly over the years. In regards to our calculations for
vaccination expenditure in the “Vaccination Rates and Vaccine Product
Prices” table, the 2023-24 flu season held the lowest in vaccine
spending compared to 2021-22 and 2022-23. Though this insight is hard to
verify given the granularity of the data, it can be inferred that this
is a result of trending lower vaccination rates.

### Implications:

Recent trends suggest that influenza uptake is on the rise, particularly
in recent years. However as concluded by our analyses, vaccination
prices do not have a significant impact on vaccine intake rates. When
analyzing vaccination rates and product prices, we calculated the total
money spent in 2021-22 on flu vaccines to be \$344,615,912, 2022-23 at
\$335,903,771, and 2023-24 at \$312,606,423. Despite this, some factors
support the potential for inaccuracy in these results. For example, the
product data might have been too granular in that it doesn't account for
factors such as the price of labor, distribution, insurance, and other
fees. Because of these gaps, it would be interesting and of great
benefit to conduct more analysis on vaccination campaigns on a national
or even global scale. An additional area for conversation would be the
correlation between vaccination rates and ED visits. Though we were able
to visualize trends in vaccination rates and ED visits independently,
the data did not overlap enough for us to merge them and extract
significantly strong results. Assuming that the merging was successful
we could provide statistical significance behind the common notion that
lower vaccination rates are correlational to higher ED visits.

### Future Analysis:

To get a more nuanced understanding of vaccination trends and their
impacts, future analyses should incorporate a greater level of
geographical granularity. By examining local trends, it would make it
possible to identify specific regions with lower vaccination rates or
higher ED visit frequencies, allowing for targeted interventions.
Further demographic analysis is also crucial to better inform outreach
efforts. A more detailed examination of specific demographic subgroups
would help pinpoint which populations are most in need of assistance,
enabling the development of tailored public health campaigns.
Additionally, a deeper contextual understanding of vaccine policy and
the product market is necessary. This would involve researching the
factors influencing vaccine pricing, availability, and public
perception. Finally, incorporating a contextual analysis that considers
the impact of COVID-19 would provide valuable insights into how the
pandemic has influenced flu vaccination behaviors and healthcare
utilization.

### Policy Recommendations:

Based on the quantitative results and observed trends, we recommend that
policy makers and organizations invest more heavily in vaccination
promotion and awareness campaigns. Increased public awareness about the
benefits of vaccination is likely to lead to higher vaccination rates,
particularly among underserved populations. While this may result in
higher government expenditures on vaccine procurement, the long-term
benefits, such as reduced emergency department (ED) visits and
associated healthcare costs, outweigh the initial investment. To
maximize the impact, these campaigns should be timed strategically,
ideally before the onset of the winter flu season, to minimize negative
health outcomes. Implementing targeted outreach programs in communities
with historically low vaccination rates could further enhance the
effectiveness of these recommendations.

### Limitations:

Several limitations impacted the scope and depth of this analysis. The
lack of overlapping data across years and seasons made it challenging to
perform robust comparisons and identify consistent trends. As previously
mentioned, the granularity of product pricing data, while being
detailed, did not provide a broader understanding of the context
surrounding vaccination campaigns and overall healthcare expenditures.
Working with and verifying cumulative totals data was difficult,
potentially affecting the accuracy of some calculated metrics. It's
important to note that while the publicly available and sourced data
used in this analysis is valuable, it could benefit from increased
comprehensiveness to allow for more in-depth research.

### References:

<https://www.cdc.gov/flu-burden/php/about/index.html#:~:text=While%20the%20burden%20of%20flu,the%20United%20States%20each%20year.&text=CDC%20estimates%20that%20flu%20has,annually%20between%202010%20and%202024.>

<https://www.cdc.gov/flu/whats-new/2023-2024-study-prevent-medical-visits.html#:~:text=December%2013%2C%202023%20%E2%80%94%20A%20new,during%20the%202022%E2%80%932023%20season>.

<https://www.cdc.gov/flu/about/season.html>
