---
title: "Final Report"
author: "Sheamin Kim"
date: "2025-03-18"
output:
  pdf_document: default
  pdf: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup_1, message=FALSE}

# Loading necessary libraries
library(tidyverse)
library(lubridate)
library(gridExtra)

```

## Background and Motivation

PASTE HERE

## Data Cleaning/Prep

```{r}
## PRICE DATA CLEANING

vax_df <- read.csv("cdc_vaccine_prices_full.csv")
inflation_df <- read.csv("inflation_cpis.csv")

#standardizing values, getting rid of $
vax_df$Private.Sector.Cost..Dose = gsub("\\$", "", vax_df$Private.Sector.Cost..Dose) 
vax_df$CDC.Cost..Dose = gsub("\\$", "", vax_df$CDC.Cost..Dose) 

vax_df$Private.Sector.Cost..Dose <- as.numeric(as.character(vax_df$Private.Sector.Cost..Dose))

vax_df$CDC.Cost..Dose <- as.numeric(as.character(vax_df$CDC.Cost..Dose))

vax_df$Date <- as.Date(vax_df$Date)


## Adding inflation data
# extract the year and convert to numeric format
vax_df$year <- as.numeric(format(vax_df$Date, "%Y"))

vax_df = merge(x = vax_df, y = inflation_df, by = "year")
vax_df$CPI <- as.numeric(as.character(vax_df$CPI))
sapply(vax_df, class)

reference_year <- 2009

# Get CPI for the reference year
reference_cpi <- vax_df$CPI[vax_df$year == reference_year]

# the type of below should be double
# print(typeof(reference_cpi)) 

# Adjust prices for inflation based on the reference CPI
vax_df$adjusted_price <- vax_df$Private.Sector.Cost..Dose * (reference_cpi / vax_df$CPI)
vax_df$adjusted_price_cdc <- vax_df$CDC.Cost..Dose * (reference_cpi / vax_df$CPI)


## RATES DATA CLEANING
df1 <- read.csv("monthly_cumulative.csv")

# Define the correct month order
month_levels <- c("SEP","OCT","NOV","DEC","JAN","FEB","MAR","APR","MAY","JUN","JUL","AUG")

# Step 1: Ensure month is a factor for proper sorting
df1 <- df1 %>%
  mutate(month = factor(month, levels = month_levels))

# getting new dose numbers
rates_df <- df1 %>%
  arrange(current_season, jurisdiction, age_group_label, month) %>%
  group_by(current_season, jurisdiction, age_group_label) %>%
  mutate(
    new_doses = numerator - lag(numerator, default = NA)  # New doses = current - previous
  ) %>%
  ungroup()

overall_pop <- rates_df %>%
  filter(age_group_label == "Overall") %>%
  select(jurisdiction, current_season, population) %>%
  rename(overall_population = population)

rates_df <- rates_df %>%
  left_join(overall_pop, by = c("jurisdiction", "current_season"))

rates_df <- rates_df %>%
  mutate(vax_rate = new_doses / coalesce(population, overall_population))

rates_df <- rates_df %>%
  mutate(start_year = as.integer(substr(current_season, 1, 4)),  # Extract the first year
         # Combine starting year with month to create a valid date
         date = as.Date(paste(start_year, month, "01", sep = "-"), format = "%Y-%b-%d"))

df_dedup <- rates_df %>%
  group_by(jurisdiction, age_group_label, current_season, date) %>%
  slice(1) %>%               # Keep only the first row for each group
  ungroup()

# 2. Sort by jurisdiction, age group, season, and date to ensure proper order for calculating new doses
df_dedup <- df_dedup %>%
  arrange(jurisdiction, age_group_label, current_season, date)

# 3. Calculate new doses by comparing the cumulative totals
df_dedup <- df_dedup %>%
  group_by(jurisdiction, age_group_label, current_season) %>%
  mutate(new_doses = numerator - lag(numerator)) %>%    # Subtract previous month from current month
  ungroup() %>%
  mutate(new_doses = ifelse(new_doses < 0, 0, new_doses))


## ED VISITS DATA CLEANING
file1 <- "ed_traj.csv"
file2 <- "ed_visits.csv"
df_1 <- read.csv(file1)
df_2 <- read.csv(file2)
# Convert week_end to Date format
df_1$week_end <- as.Date(df_1$week_end, format="%Y-%m-%d")
df_2$week_end <- as.Date(df_2$week_end, format="%Y-%m-%d")

# Clean df1 (Trajectories dataset) - Select relevant columns
df1_clean <- df_1 %>%
  select(week_end, geography, county, percent_visits_influenza) %>%
  filter(!is.na(percent_visits_influenza))
# Clean df2 (Demographics dataset) - Select flu data only
df2_clean <- df_2 %>%
  filter(pathogen == "Influenza") %>% # Select only Influenza-related ED visits
  select(week_end, geography, percent_visits) %>%
  rename(percent_visits_influenza = percent_visits) %>%
  filter(!is.na(percent_visits_influenza))

# Merge both datasets for better insights
df_combined <- bind_rows(df1_clean, df2_clean)

df_combined$Date <- as.Date(df_combined$week_end)

df_combined$year <- format(df_combined$week_end, "%Y")
df_combined$month <- format(df_combined$week_end, "%m")
df_combined$month_abbr <- month.abb[as.numeric(df_combined$month)]

seasonal <- df_combined %>%
  filter(county == "All") %>%
  group_by(Date) %>%
  summarise(percent_visits_influenza = mean(percent_visits_influenza))

seasonal$Date <- as.Date(seasonal$Date)
seasonal$year <- format(seasonal$Date, "%Y")
seasonal$month <- format(seasonal$Date, "%m")
seasonal$month_abbr <- month.abb[as.numeric(seasonal$month)]

```

## Datasets

1.  Flu vaccination rates

-   <https://healthdata.gov/dataset/Monthly-Cumulative-Number-and-Percent-of-Persons-W/8y48-wjrp/about_data> 

The flu vaccination rates dataset, sourced from HealthData.gov and
maintained by the CDC, provides monthly cumulative counts and
percentages of individuals who have received at least one dose of the
influenza vaccine. The data spans multiple flu seasons, from 2019 to
2023, and is categorized by age group and jurisdiction (states,
territories, and select cities). The data set is compiled from
Immunization Information Systems (IIS), which aggregate vaccine
administration data from various public health agencies.

This data set offers insights into vaccination trends over time and
across different demographic groups. The cumulative nature of the
records ensures that historical data is preserved, allowing for trend
analysis. However, the data set has limitations, including variations in
data completeness across jurisdictions and differences in state policies
regarding vaccine data reporting. The population denominators used for
calculating vaccination rates are sourced from the U.S. Census Bureau's
2020 estimates. Standard errors are not provided, as the data includes
all vaccinations rather than a sample.

[EXPLANATION FOR THESE HERE]

```{r}
## Summary Table
monthly_trends <- rates_df %>%
  group_by(current_season, month) %>%
  summarise(new_doses = sum(new_doses, na.rm = TRUE)) %>%
  arrange(current_season, month)

print(monthly_trends)
```

```{r}
## Basic Bar Chart
df_clean <- df1 %>%
  filter(!is.na(numerator)) %>% # Remove rows where numerator is NA
  arrange(jurisdiction, current_season, age_group_label, month) %>% # Sort by jurisdiction, season, and month
  group_by(jurisdiction, current_season, age_group_label) %>% # Group by jurisdiction, season, and age group
  mutate(monthly_doses = numerator - lag(numerator)) %>% # Subtract previous month from current to get actual doses
  ungroup() %>% # Remove the grouping
  filter(!is.na(monthly_doses) & monthly_doses >= 0) # Remove NAs and negative values (in case of any

population_per_season_jurisdiction <- df_clean %>%
  filter(age_group_label == "Overall") %>% # Only include rows where age_group_label is "Overall"
  group_by(current_season, jurisdiction) %>%
  summarise(
    unique_population = unique(population), # Get a single unique population value per jurisdiction
    total_doses = sum(monthly_doses, na.rm = TRUE)) %>%
  group_by(current_season) %>%
  summarise(
    total_population = sum(unique_population, na.rm = TRUE), # Sum up the unique population across all jurisdictions
    total_doses = sum(total_doses, na.rm = TRUE)) %>%
  mutate(vaccination_rate = (total_doses / total_population) * 2.5)

population_per_season_jurisdiction %>%
  ggplot(aes(x = current_season, y = vaccination_rate, fill = current_season)) +
  geom_col() +
  geom_text(aes(label = scales::percent(vaccination_rate)), vjust = -0.5, size = 4) +
  labs(title = "Flu Vaccination Rate by Season",
       x = "Flu Season",
       y = "Vaccination Rate") +
  theme_minimal() +
  theme(legend.position = "none")

```

2.  Flu vaccination expenditure

-   <https://www.cdc.gov/vaccines-for-children/php/awardees/current-cdc-vaccine-price-list.html>

-   <https://www.cdc.gov/vaccines/programs/vfc/awardees/vaccine-management/price-list/archive.html> 

The flu vaccination expenditure dataset is derived from CDC’s Vaccine
Price Lists, which detail both public-sector contract prices and
private-sector prices for influenza vaccines. The dataset includes
pricing information for pediatric and adult flu vaccines, with
historical records dating back to 2001. The primary source for current
vaccine prices is the CDC’s publicly available vaccine price list, while
archived prices are stored separately.

The dataset includes details such as vaccine brand names, National Drug
Codes (NDCs), packaging information, CDC cost per dose, private sector
cost per dose, contract end dates, and manufacturers. This data allows
for an analysis of pricing trends over time, identifying fluctuations in
vaccine costs and potential disparities between public and private
sector pricing. However, obtaining historical data required web scraping
or API access, as the archived prices are distributed across multiple
web pages.

[EXPLANATION HERE] 

```{r}
## Summary Table
summary_table <- vax_df %>%
  group_by(year) %>%
  summarise(
    num_products = n(),
    avg_cdc_price = mean(CDC.Cost..Dose, na.rm = TRUE),
    avg_private_price = mean(Private.Sector.Cost..Dose, na.rm = TRUE),
    avg_adj_cdc_price = mean(adjusted_price_cdc, na.rm = TRUE),
    avg_adj_private_price = mean(adjusted_price, na.rm = TRUE),
    min_private_price = min(Private.Sector.Cost..Dose, na.rm = TRUE),
    max_private_price = max(Private.Sector.Cost..Dose, na.rm = TRUE),
  )

print(summary_table)
```

3.  Flu emergency department visit rates

-   <https://healthdata.gov/dataset/NSSP-Emergency-Department-Visit-Trajectories-by-St/hr4c-e7p6/about_data> 

-   <https://healthdata.gov/dataset/NSSP-Emergency-Department-Visits-COVID-19-Flu-RSV-/vfw5-fbqw/about_data>

The flu emergency department (ED) visit rates dataset is sourced from
the National Syndromic Surveillance Program (NSSP) and published on
HealthData.gov. This dataset provides the percentage of emergency
department visits that are attributed to influenza, alongside data for
other respiratory illnesses such as COVID-19 and RSV. The dataset spans
from 2022 to 2025 and is updated weekly.

The data set is available in two formats:

-   **NSSP Emergency Department Visit Trajectories by State and
    Sub-State Regions**: This data set reports the percentage of ED
    visits for flu at both state and sub-state (Health Service Area)
    levels. It also includes trend classifications (increasing,
    decreasing, or stable) based on statistical models.

-   **NSSP Emergency Department Visits by Demographic Category**: This
    dataset categorizes ED visits for influenza by demographic variables
    such as age, sex, and race/ethnicity. It provides insights into
    disparities in flu-related ED visits across different population
    groups.

The data is collected from health facilities participating in the NSSP
and is intended to track trends over time.

[EXPLANATION HERE]

```{r}
# Create yearly summary table (limit to top 10 states)
summary_table <- df_combined %>%
  mutate(year = year(week_end)) %>%
  group_by(year, geography) %>%
  summarize(avg_percent_influenza = mean(percent_visits_influenza, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(avg_percent_influenza)) %>%
  group_by(year) %>%
  slice_max(order_by = avg_percent_influenza, n = 10) # Keep only the top 10 states

print(summary_table)
```

## Methods

TEXT HERE

## Results

### Individual Analysis

#### Price Data

```{r}
## EDIT TITLE
plot_1 <- vax_df %>%
  group_by(year) %>%
  summarise(average_price = mean(Private.Sector.Cost..Dose), 
            average_adjust_price = mean(adjusted_price)) %>%
  pivot_longer(cols = c("average_price", "average_adjust_price"), 
               names_to = "price_type", 
               values_to = "price") %>%
  ggplot(aes(x=factor(year), y=price, fill=price_type)) +
  geom_col(position="dodge") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(labels = 2009:2025, breaks = 2009:2025) +
  labs(title = "Average Price of 10 Influenza Vaccine Doses",
       x = "Year",
       y = "Price",
       fill = "Adjusted for Inflation") 

print(plot_1)
```

EXPLAIN HERE

```{r}

##EDIT TITLES AND STUFF
adj_private_cdc_comparison_plot <- vax_df %>%
  group_by(year) %>%
  summarise(average_priv_price = mean(adjusted_price), 
            average_cdc_price = mean(adjusted_price_cdc)) %>%
  pivot_longer(cols = c("average_priv_price", "average_cdc_price"), 
               names_to = "price_type", 
               values_to = "price") %>%
  ggplot(aes(x=factor(year), y=price, group=price_type)) +
  geom_line(aes(color=price_type))

adj_private_cdc_comparison_plot


private_cdc_comparison_plot <- vax_df %>%
  group_by(year) %>%
  summarise(average_priv_price = mean(Private.Sector.Cost..Dose), 
            average_cdc_price = mean(CDC.Cost..Dose)) %>%
  pivot_longer(cols = c("average_priv_price", "average_cdc_price"), 
               names_to = "price_type", 
               values_to = "price") %>%
  ggplot(aes(x=factor(year), y=price, group=price_type)) +
  geom_line(aes(color=price_type))

private_cdc_comparison_plot

print(grid.arrange(private_cdc_comparison_plot, adj_private_cdc_comparison_plot))
```

EXPLAIN ABOVE PLOT HERE

```{r}
test <- t.test(vax_df$Private.Sector.Cost..Dose, vax_df$CDC.Cost..Dose,
               alternative = "greater")
test
```

EXPLAIN T-TEST RESULTS HERE

```{r}
# lin reg - is there a correlation between year and vaccine price
model <- lm(adjusted_price ~ year, data = vax_df)
print(summary(model))

ggplot(vax_df, aes(x = year, y = adjusted_price)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Linear Regression for Year and Adjusted Vaccine Prices",
       x = "Year",
       y = "Adjusted Private Sector Price (for Inflation)")
```

EXPLAIN PLOT HERE

#### Vaccination Rates

```{r}
## ANOVA CODE HERE
```

EXPLAIN OUTPUT HERE

#### Emergency Department Visits

```{r}
## MAKE THIS PRETTIER
df_for_plot <- df_combined %>%
  group_by(Date) %>%
  summarise(value = mean(percent_visits_influenza))

plot_time <- ggplot(data = df_for_plot) +
  geom_line(aes(x = Date, y = value))

plot_time
```

EXPLAIN PLOT HERE

```{r}
anova_result <- aov(percent_visits_influenza ~ as.factor(year), data = df_combined)
summary(anova_result) # Print ANOVA test result
```

EXPLAIN FINDINGS

### Relationship Analysis

#### Vaccination Rates and Emergency Department Visits

```{r}
df_for_cop <- df_dedup %>%
  group_by(date) %>%
  summarise(plot_col = sum(new_doses, na.rm = TRUE))

## MAKE THIS PRETTIER
ggplot(seasonal, aes(x = factor(month_abbr), y = percent_visits_influenza, color = as.factor(year), group = year)) +
  geom_smooth()

#ggplot() +
 # geom_line(data=seasonal, aes(x = Date, y = percent_visits_influenza)) +
  #geom_line(data=temp_df, aes(x = date, y = avg_rate))

merged = merge(df_for_cop, seasonal, by.x="date", by.y="Date")

## BAD TEST HERE
# cor.test(merged$plot_col, merged$percent_visits_influenza, method = "pearson")

```

text here explaining

#### Vaccination Rates and Vaccine Product Prices

```{r}

all_totals <- df_dedup %>%
  group_by(date) %>%
  summarise(total_doses = sum(new_doses, na.rm = TRUE))

all_totals$year <- format(all_totals$date, "%Y")

all_totals$doses_div_ten <- (all_totals$total_doses) / 10

yr_totals <- all_totals %>%
  group_by(year) %>%
  summarise(total_doses = sum(total_doses))


price_table <- vax_df %>%
  group_by(year) %>%
  summarise(cost = mean(adjusted_price))

price_table$cost_per_dose <- (price_table$cost) / 10

price_yr_totals <- merge(x = yr_totals, y = price_table, by = "year")

price_yr_totals$money_spent <- price_yr_totals$total_doses * price_yr_totals$cost_per_dose

print(price_yr_totals)
```

SOME TEXT HERE

```{r}
anova_result <- aov(money_spent ~ as.factor(year), data = price_yr_totals)
summary(anova_result)
```

EXPLANATION HERE

## Discussion

text here

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to
prevent printing of the R code that generated the plot.
